\bf Eye Fitting Straight Lines in the Modern Era

### 2021-09-07 Eye-Fitting-Stright-Lines-in-the-Modern-Era.tex
updated bib. began bullet points.

  \author{
        Emily A. Robinson 1 \thanks{The authors gratefully acknowledge
\ldots{}} \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Susan VanderPlas 2 \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Reka Howard 3 \\
    Department of Statistics, University of Nebraska - Lincoln\\
      }
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Fitting lines by eye through a set of points has been explored since the
20th century. Common methods of fitting trends by eye involve
maneuvering a string, black thread, or ruler until the fit is suitable,
then drawing the line through the set of points. In 2015, the New York
Times introduced an interactive feature, called `You Draw It'. Readers
are asked to input their own assumptions about various metrics and
compare how these assumptions relate to reality. The New York Times team
utilizes Data Driven Documents (D3) that allows readers to predict these
metrics through the use of drawing a line on their computer screen with
their computer mouse. In my research, I created `You Draw It'
interactive graphics using the r2d3 package in R statistical software
and validated this as a tool and method for testing graphics. I
recruited participants via crowdsourcing websites and replicated the
study found in Eye Fitting Straight Lines (Mosteller et al., 1981) in
order to establish you draw it as a tool and method for testing
graphics. Participats were directed to an RShiny application link and
shown points following a linear trend and asked to draw a line through
the data points using their computer mouse. A total of 39 participants
completed trendlines for 256 unique interactive plots. A linear mixed
model and a generalized additive mixed model were used to determine the
relationship between the participants drawn points and the best fit
lines. Results from my study were consistent with those found in the
previous study. When shown points following a linear trend, participants
tended to fit the slope of the first principal component over the slope
of the least squares regression line. This trend was most prominent when
shown data simulated with larger variances. The reproducibility of these
results serve as evidence of the reliability of the you draw it method.
Future work is necessarry to implement the `You Draw It' tool as a
method of testing graphics. {[}200 word limit{]}
\end{abstract}

\noindent%
{\it Keywords:} Graphics, Regression, Perception
\vfill

@@

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points
\citep{finney_subjective_1951, mosteller_eye_1981}. Common methods of
fitting trends by eye involve maneuvering a string, black thread, or
ruler until the fit is suitable, then drawing the line through the set
of points. In \citet{finney_subjective_1951}, it was of interest to
determine the effect of stopping iterative maximum likelihood
calculations after one iteration. Many techniques in statistical
analysis are performed with the aid of iterative calculations such as
Newton's method or Fisher's scoring. Guesses are made at the best
estimates of certain parameters and these guesses are then used as the
basis of a computation which yields a new set of approximation to the
parameter estimates; this same procedure is the performed on the new
parameter estimates and the computing cycle is repeated until
convergence, as determined by the statistician, is reached. The author
was interested in whether one iteration of calculations was sufficient
in the estimation of parameters connected with dose-response
relationships. One measure of interest is the relative potency between a
test preparation of doses and standard preparation of does; relative
potency is calculated as the ratio of two equally effective doses
between the two preparation methods. \cref{fig:subjective-judgement}
shows a pair of parallel probit responses in a biological assay. The
x-axis is the \(\log_{1.5}\) dose level for four dose levels (for
example, doses 4, 6, 9, and 13 correspond correspond to equally spaced
values on a logarithmic scale, labeled 0, 1, 2, and 3) and the y-axis is
the corresponding probit response as calculated in
\citet{finney1948table}; circles correspond to the test preparation
method while the crosses correspond to the standard preparation method.
For these sort of assays, the does-response relationship follows a
linear regression of the probit response on the logarithm of the dose
levels; the two preparation methods can be constrained to be parallel
\citep{jerne1949validity}, limiting the relative potency to one
consistent value. In this study, twenty-one scientists were recruited
via postal mail and asked to ``rule two lines'' in order to judge by eye
the positions for a pair of parallel probit regression lines in a
biological assay (\cref{fig:subjective-judgement}). The author then
computed one iterative calculation of the relative potency based on
starting values as indicated by the pair of lines provided by each
participant and compared these relative potency estimates to that which
was estimated by the full probit technique (reaching convergence through
multiple iterations). Results indicated that one cycle of iterations for
calculating the relative potency was sufficient based on the starting
values provided by eye from the participants.

{\centering \includegraphics[width=0.5\linewidth,]{images/subjective-judgement-plot}

\caption{Subjective Judgement in Statistical Analysis (1951) Parallel Probits}\label{fig:subjective-judgement}
\end{figure}

Thirty years later, Mosteller et al.~(1981), sought to understand the
properties of least squares and other computed lines by establishing one
systematic method of fitting lines by eye. The authors recruited 153
graduate students and post doctoral researchers in Introductory
Biostatistics. Participants were asked to fit lines by eye to four sets
of points (\cref{fig:mosteller-eyefitting-plot}) using an 8.5 x 11 inch
transparency with a straight line etched completely across the middle. A
latin square design \citep{giesbrecht2004planning} with packets of the
set of points stapled together in four different sequences was used to
determine if there is an effect of order of presentation. It was found
that order of presentation had no effect and that participants tended to
fit the slope of the first principal component (error minimized
orthogonally, both horizontal and vertical, to the regression line) over
the slope of the least squares regression line (error minimized
vertically to the regression line).

{\centering \includegraphics[width=0.7\linewidth,]{images/eyefitting-straight-lines-plots}

\caption{Eye Fitting Straight Lines (1981) Data Sets}\label{fig:mosteller-eyefitting-plot}
\end{figure}

In 2015, the New York Times introduced an interactive feature, called
You Draw It
\citep{aisch_cox_quealy_2015, buchanan_park_pearce_2017, katz_2017}.
Readers are asked to input their own assumptions about various metrics
and compare how these assumptions relate to reality. The New York Times
team utilizes Data Driven Documents (D3) that allows readers to predict
these metrics through the use of drawing a line on their computer screen
with their computer mouse. \cref{fig:nyt-caraccidents} \citep{katz_2017}
is one such example in which readers are asked to draw the line for the
missing years providing what they estimate to be the number of Americans
who have died every year from car accidents, since 1990. After the
reader has completed drawing the line, the actual observed values are
revealed and the reader may check their estimated knowledge against the
actual reported data.

{\centering \includegraphics[width=0.75\linewidth,]{images/nyt-caraccidents-frame4}

\caption{New York Times You Draw It Feature}\label{fig:nyt-caraccidents}
\end{figure}

@@

The goal of this paper is to establish you draw it as a tool for
measuring predictions of trends fitted by eye and a method for testing
graphics, the second then applies you draw it to test an individual's
ability to make predictions for exponentially increasing data. In order
to validate you draw it as a method for testing graphics, the first
sub-study, referred to as Eye Fitting Straight Lines in the Modern Era,
replicated the experiment and results found in Mosteller et al.~(1981).

@@

In addition to fitting trend lines over the residuals between the
participant drawn values and fitted values, an OLS sum of squares and
PCA sums of squares measure was calculated for each you draw it plot.
Sums of squares between the two fits were compared using the lmer
function in the lme4 package in R \citep{lme4} to run a linear mixed
model (LMM) with a log transformation. Parameter choice (S, F, V, N),
fit (OLS, PCA), and the interaction between parameter choice and fit
were treated as fixed effects with a random participant effect. Define
\(SS_{ijk}\) as the sums of squares for parameter choice
\(i = 1,2,3,4\), fit \(j=1,2\), and participant
\(k = 1,...,N_{participant}\). The LMM equation is given by:
\begin{equation}
\log\left(SS_{ijk}\right) = \alpha_i + \beta_j + \alpha\beta_{ij} + p_{j} + \epsilon_{ijk}
\end{equation}

\begin{itemize}
\tightlist
\item
  \(\alpha_i\) denotes the effect of the \(i^{th}\) parameter choice
\item
  \(\beta_j\) denotes the effect of the \(j^{th}\) fit
\item
  \(\alpha\beta_{ij}\) denotes the interaction between the \(i^{th}\)
  parameter choice and \(j^{th}\) fit
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
  the \(k^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}

The estimated odds ratios between the OLS fit and PCA fit for each
parameter choice are displayed in \cref{fig:eyefitting-ss-oddsratio}.
While there is no significant effect of fit for any parameter choices,
there is indication of the trend found previously in the residual
regression models indicating that participants tend to fit the slope of
the first principal component over the slope of the ordinary least
squares regression line, particularly for trends with large variation
(parameter choices N and F).

{\centering \includegraphics[width=0.75\linewidth,]{Eye-Fitting-Stright-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-ss-oddsratio-1}

\caption{Eye Fitting Straight Lines in the Modern Era Sum of Squares Results}\label{fig:eyefitting-ss-oddsratio}
\end{figure}

@@

The intent of this paper was to establish you draw it as a tool for
testing graphics then use this tool to determine the cognitive
implications of displaying data on the log scale. Eye Fitting Straight
Lines in the Modern Era replicated the results found in Mosteller et
al.~(1981). When shown points following a linear trend, participants
tended to fit the slope of the first principal component over the slope
of the least squares regression line. This trend was most prominent when
shown data simulated with larger variances. The reproducibility of these
results serve as evidence of the reliability of the you draw it method.

>>>>

  \author{
        Emily A. Robinson 1 \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Susan VanderPlas 2 \\
    Department of Statistics, University of Nebraska - Lincoln\\
     and \\     Reka Howard 3 \\
    Department of Statistics, University of Nebraska - Lincoln\\
      }
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Fitting lines by eye through a set of points has been explored since the
20th century. Common methods of fitting trends by eye involve
maneuvering a string, black thread, or ruler until the fit is suitable,
then drawing the line through the set of points. In 2015, the New York
Times introduced an interactive feature, called `You Draw It'. Readers
are asked to input their own assumptions about various metrics and
compare how these assumptions relate to reality. The New York Times team
utilizes Data Driven Documents (D3) that allows readers to predict these
metrics by drawing a line on their computer screen with their computer
mouse. In my research, I established `You Draw It' as a method for
graphical testing by adapting the New York Times feature. I recruited
participants via crowdsourcing websites and replicated the study found
in Eye Fitting Straight Lines (Mosteller et al., 1981). Participants
were directed to an RShiny application link and shown points following a
linear trend and asked to draw a line through the data points using
their computer mouse; task plots were generated using the r2d3 package
in R statistical software. Results from my study were consistent with
those found in the previous study; when shown points following a linear
trend, participants tended to fit the slope of the first principal
component over the slope of the least-squares regression line. This
trend was most prominent when shown data simulated with larger
variances. The reproducibility of these results serves as evidence of
the reliability of the you draw it method. Future work is necessary to
implement the `You Draw It' tool as a method of testing graphics. {[}200
word limit{]}
\end{abstract}

\noindent%
{\it Keywords:} Graphics, Regression, Graph
Perception, Scatterplot, Cognitive Bias
\vfill

@@

Advanced technology and computing power have promoted data visualization
as a central tool in modern data science. \citet{unwin2020data} defines
data visualization as the art of drawing graphical charts in order to
display data. Graphics are useful for data cleaning, exploring data
structure, and have been an essential component in communicating
information for the last 200 years \citep{lewandowsky1989perception}.
Although statistical graphics have become widely used and valued in
science, business, and in many other aspects of life, as creators of
graphics, we are too accepting of them as default without asking
critical questions about the graphics we create or view (Unwin, 2020).

\hypertarget{graph-perception}{%
\subsection{Graph Perception}\label{graph-perception}}

\hypertarget{testing-statistical-graphics}{%
\subsection{Testing Statistical
Graphics}\label{testing-statistical-graphics}}

One way in which we evaluate the effectiveness of charts is through the
use of graphical tests. We could ask participants to identify
differences in graphs, read information off of a chart accurately, use
data to make correct real-world decisions, or predict the next few
observations. All of these types of tests require different levels of
use and manipulation of the information being presented in the chart.

\begin{itemize}
\tightlist
\item
  Psychophysics (speed and accuracy)
\item
  Lineups
\end{itemize}

\hypertarget{trend-judgement}{%
\subsection{Trend Judgement}\label{trend-judgement}}

\begin{itemize}
\tightlist
\item
  \citet{finney1951subjective}
\item
  \citet{mosteller_eye_1981}
\item
  \citet{ciccione2021can}
\end{itemize}

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points
\citep{finney1951subjective, mosteller_eye_1981}. Common methods of
fitting trends by eye involve maneuvering a string, black thread, or
ruler until the fit is suitable, then drawing the line through the set
of points. In \citet{finney1951subjective}, it was of interest to
determine the effect of stopping iterative maximum likelihood
calculations after one iteration. Many techniques in statistical
analysis are performed with the aid of iterative calculations such as
Newton's method or Fisher's scoring. Guesses are made at the best
estimates of certain parameters and these guesses are then used as the
basis of a computation which yields a new set of approximation to the
parameter estimates; this same procedure is the performed on the new
parameter estimates and the computing cycle is repeated until
convergence, as determined by the statistician, is reached. The author
was interested in whether one iteration of calculations was sufficient
in the estimation of parameters connected with dose-response
relationships. One measure of interest is the relative potency between a
test preparation of doses and standard preparation of does; relative
potency is calculated as the ratio of two equally effective doses
between the two preparation methods. \cref{fig:subjective-judgement}
shows a pair of parallel probit responses in a biological assay. The
x-axis is the \(\log_{1.5}\) dose level for four dose levels (for
example, doses 4, 6, 9, and 13 correspond correspond to equally spaced
values on a logarithmic scale, labeled 0, 1, 2, and 3) and the y-axis is
the corresponding probit response as calculated in
\citet{finney1948table}; circles correspond to the test preparation
method while the crosses correspond to the standard preparation method.
For these sort of assays, the does-response relationship follows a
linear regression of the probit response on the logarithm of the dose
levels; the two preparation methods can be constrained to be parallel
\citep{jerne1949validity}, limiting the relative potency to one
consistent value. In this study, twenty-one scientists were recruited
via postal mail and asked to ``rule two lines'' in order to judge by eye
the positions for a pair of parallel probit regression lines in a
biological assay. The author then computed one iterative calculation of
the relative potency based on starting values as indicated by the pair
of lines provided by each participant and compared these relative
potency estimates to that which was estimated by the full probit
technique (reaching convergence through multiple iterations). Results
indicated that one cycle of iterations for calculating the relative
potency was sufficient based on the starting values provided by eye from
the participants.

Thirty years later, Mosteller et al.~(1981), sought to understand the
properties of least squares and other computed lines by establishing one
systematic method of fitting lines by eye. The authors recruited 153
graduate students and post doctoral researchers in Introductory
Biostatistics. Participants were asked to fit lines by eye to four
scatterplots using an 8.5 x 11 inch transparency with a straight line
etched completely across the middle. A latin square design
\citep{giesbrecht2004planning} with packets of the set of points stapled
together in four different sequences was used to determine if there is
an effect of order of presentation. It was found that order of
presentation had no effect and that participants tended to fit the slope
of the first principal component (error minimized orthogonally, both
horizontal and vertical, to the regression line) over the slope of the
least squares regression line (error minimized vertically to the
regression line).

In 2015, the New York Times introduced an interactive feature, called
You Draw It
\citep{aisch_cox_quealy_2015, buchanan_park_pearce_2017, katz_2017}.
Readers are asked to input their own assumptions about various metrics
and compare how these assumptions relate to reality. The New York Times
team utilizes Data Driven Documents (D3) that allows readers to predict
these metrics through the use of drawing a line on their computer screen
with their computer mouse. \citep{katz_2017} is one such example in
which readers are asked to draw the line for the missing years providing
what they estimate to be the number of Americans who have died every
year from car accidents, since 1990. After the reader has completed
drawing the line, the actual observed values are revealed and the reader
may check their estimated knowledge against the actual reported data.

@@

\hypertarget{research-objective}{%
\subsection{Research objective}\label{research-objective}}

The goal of this paper is to establish you draw it as a tool for
measuring predictions of trends fitted by eye and a method for testing
graphics. In order to validate you draw it as a method for testing
graphics, the first sub-study, referred to as Eye Fitting Straight Lines
in the Modern Era, replicated the experiment and results found in
Mosteller et al.~(1981).

@@

@@

The intent of this paper was to establish you draw it as a tool. Eye
Fitting Straight Lines in the Modern Era replicated the results found in
Mosteller et al.~(1981). When shown points following a linear trend,
participants tended to fit the slope of the first principal component
over the slope of the least squares regression line. This trend was most
prominent when shown data simulated with larger variances. The
reproducibility of these results serve as evidence of the reliability of
the you draw it method.

\hypertarget{future-work}{%
\section{Future Work}\label{future-work}}

Use the method. \emph{Should probably elaborate here..}
